Data analysis code

### 2024-04-19 code/data_analysis_code.qmd
Project Part 5
Completed part 5 of the project (might review and do a few edits later)

---
title: "Project Part 3"
subtitle: "Data analysis code"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format: html
editor: visual
---

@@

```{r}
library(here)
library(readxl)
library(tidyverse)
library(corrplot)
library(RColorBrewer)
library(ggpubr)
library(rstatix)
library(tidymodels)
```

@@

To answer question 1, I will plot the correlations in a correlation plot using the `corrplot()` function. I used the previous code, since it was already developed, but editing to add more detail to the plot.

@@

To answer question 2, I will conduct a T-test using the `tidymodels` package

@@

  tibble( #Format the tibble
    variable = variable,
    statistic = t_test$statistic,
    p_value = t_test$p.value,
    estimate = t_test$estimate,
    null_value = t_test$null.value,
    alternative = t_test$alternative,
    method = t_test$method,
    conf_low = t_test$conf.int[1],
    conf_high = t_test$conf.int[2]
  )
})

@@

To answer question 3, I will conduct a regression analysis. But first I will merge the data from `hapin_filters` with the `concentration` data. Modeling can be difficult if there are missing values, so I will additionally clean the `hapin_merged` df, so it removes all NAs.

@@

#Remove pm25, bc and co, and additionally drop all NAs
hapin_merged <- hapin_merged %>%
  select(-c(pm25, bc, co)) %>%
  drop_na()

@@

Now conducting the model selection using the `tidymodels` package. For the models, I used the gamma distribution family, since the concentrations of each chemical species could be explained by the gamma distribution. I used all the recorded exposures except incence and generator, since these had low variability in the answers.

```{r}
# Create a generalized linear regression model specification using gamma distribution (This will be used for all the models)
gamma_spec <- linear_reg(mode = "regression") %>%
  set_engine("glm", family = Gamma(link = "log"))

###########To predict the Al variable##############
# Create the recipe for preprocessing
recipeAl <- recipe(Al ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitAl <- workflow() %>%
  add_recipe(recipeAl) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsAl <- fitAl %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsAl)

###########To predict the BC variable##############
# Create the recipe for preprocessing
recipeBC <- recipe(BC ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitBC <- workflow() %>%
  add_recipe(recipeBC) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsBC <- fitBC %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsBC)

###########To predict the Ca variable##############
# Create the recipe for preprocessing
recipeCa <- recipe(Ca ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitCa <- workflow() %>%
  add_recipe(recipeCa) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsCa <- fitCa %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsCa)

###########To predict the Fe variable##############
# Create the recipe for preprocessing
recipeFe <- recipe(Fe ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitFe <- workflow() %>%
  add_recipe(recipeFe) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsFe <- fitFe %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsFe)

###########To predict the K variable##############
# Create the recipe for preprocessing
recipeK <- recipe(K ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitK <- workflow() %>%
  add_recipe(recipeK) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsK <- fitK %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsK)

###########To predict the Mg variable##############
# Create the recipe for preprocessing
recipeMg <- recipe(Mg ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitMg <- workflow() %>%
  add_recipe(recipeMg) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMg <- fitMg %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMg)

###########To predict the Mn variable##############
# Create the recipe for preprocessing
recipeMn <- recipe(Mn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitMn <- workflow() %>%
  add_recipe(recipeMn) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMn <- fitMn %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMn)

###########To predict the S variable##############
# Create the recipe for preprocessing
recipeSul <- recipe(S ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitSul <- workflow() %>%
  add_recipe(recipeSul) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsSul <- fitSul %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsSul)

###########To predict the Si variable##############
# Create the recipe for preprocessing
recipeSi <- recipe(Si ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitSi <- workflow() %>%
  add_recipe(recipeSi) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsSi <- fitSi %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsSi)

###########To predict the Ti variable##############
# Create the recipe for preprocessing
recipeTi <- recipe(Ti ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitTi <- workflow() %>%
  add_recipe(recipeTi) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsTi <- fitTi %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsTi)

###########To predict the Zn variable##############
# Create the recipe for preprocessing
recipeZn <- recipe(Zn ~ stove + smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitZn <- workflow() %>%
  add_recipe(recipeZn) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsZn <- fitZn %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsZn)
```

Based on the previous models, it seems like the most important variables are `fueltype` (as expected), `kerosene`, `stove_other` and `smoke` depending on the chemical species. I was worried of the low variability on responses for some of the variables so I am going to use only the variables `micro_exposures`, `macro_exposures`, `kerosene`, `stove_other` and `fueltype`.

```{r}
###########To predict the Al variable##############
# Create the recipe for preprocessing
recipeAl2 <- recipe(Al ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitAl2 <- workflow() %>%
  add_recipe(recipeAl2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsAl2 <- fitAl2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsAl2)

###########To predict the BC variable##############
# Create the recipe for preprocessing
recipeBC2 <- recipe(BC ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitBC2 <- workflow() %>%
  add_recipe(recipeBC2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsBC2 <- fitBC2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsBC2)

###########To predict the Ca variable##############
# Create the recipe for preprocessing
recipeCa2 <- recipe(Ca ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitCa2 <- workflow() %>%
  add_recipe(recipeCa2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsCa2 <- fitCa2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsCa2)

###########To predict the Fe variable##############
# Create the recipe for preprocessing
recipeFe2 <- recipe(Fe ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitFe2 <- workflow() %>%
  add_recipe(recipeFe2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsFe2 <- fitFe2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsFe2)

###########To predict the K variable##############
# Create the recipe for preprocessing
recipeK2 <- recipe(K ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitK2 <- workflow() %>%
  add_recipe(recipeK2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsK2 <- fitK2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsK2)

###########To predict the Mg variable##############
# Create the recipe for preprocessing
recipeMg2 <- recipe(Mg ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitMg2 <- workflow() %>%
  add_recipe(recipeMg2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMg2 <- fitMg2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMg2)

###########To predict the Mn variable##############
# Create the recipe for preprocessing
recipeMn2 <- recipe(Mn ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitMn2 <- workflow() %>%
  add_recipe(recipeMn2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsMn2 <- fitMn2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsMn2)

###########To predict the S variable##############
# Create the recipe for preprocessing
recipeSul2 <- recipe(S ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitSul2 <- workflow() %>%
  add_recipe(recipeSul2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsSul2 <- fitSul2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsSul2)

###########To predict the Si variable##############
# Create the recipe for preprocessing
recipeSi2 <- recipe(Si ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitSi2 <- workflow() %>%
  add_recipe(recipeSi2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsSi2 <- fitSi2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsSi2)

###########To predict the Ti variable##############
# Create the recipe for preprocessing
recipeTi2 <- recipe(Ti ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitTi2 <- workflow() %>%
  add_recipe(recipeTi2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsTi2 <- fitTi2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsTi2)

###########To predict the Zn variable##############
# Create the recipe for preprocessing
recipeZn2 <- recipe(Zn ~ micro_exposures + macro_exposures + stove_other + fueltype, data = hapin_merged) %>%
  step_dummy(all_nominal(), -all_outcomes())

#Fit the model
fitZn2 <- workflow() %>%
  add_recipe(recipeZn2) %>%
  add_model(gamma_spec) %>%
  fit(data = hapin_merged)

# Extract and print the results of the model
resultsZn2 <- fitZn2 %>%
  pull_workflow_fit() %>%
  tidy()

print(resultsZn2)
```

Using this approach it seems like only `fueltype` and `other_stove` are almost all of the variables important to make predictions of the chemical species.

And now evaluating both models for all the species. First I will set the resampling method using cross-validation.

```{r}
#Set the seed for reproducibility
set.seed(1234)

# Set up resampling method with 10 folds cross-validation
samples <- vfold_cv(hapin_merged, v = 10)
```

And now evaluating both models of `Al` using the RMSE and R-squared values for both models. The same process will be described below for the other chemical species.

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Al_evaluation <- workflow() %>%
  add_recipe(recipeAl) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Al_metrics <- Al_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Al2_evaluation <- workflow() %>%
  add_recipe(recipeAl2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Al2_metrics <- Al2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(Al_metrics)
print(Al2_metrics)
```

Now for BC...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
BC_evaluation <- workflow() %>%
  add_recipe(recipeBC) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
BC_metrics <- BC_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
BC2_evaluation <- workflow() %>%
  add_recipe(recipeBC2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
BC2_metrics <- BC2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(BC_metrics)
print(BC2_metrics)
```

Now for Ca...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Ca_evaluation <- workflow() %>%
  add_recipe(recipeCa) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Ca_metrics <- Ca_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Ca2_evaluation <- workflow() %>%
  add_recipe(recipeCa2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Ca2_metrics <- Ca2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(Ca_metrics)
print(Ca2_metrics)
```

Now for Fe...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Fe_evaluation <- workflow() %>%
  add_recipe(recipeFe) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Fe_metrics <- Fe_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Fe2_evaluation <- workflow() %>%
  add_recipe(recipeFe2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Fe2_metrics <- Fe2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(Fe_metrics)
print(Fe2_metrics)
```

Now for K...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
K_evaluation <- workflow() %>%
  add_recipe(recipeK) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
K_metrics <- K_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
K2_evaluation <- workflow() %>%
  add_recipe(recipeK2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
K2_metrics <- K2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(K_metrics)
print(K2_metrics)
```

Now for Mg...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Mg_evaluation <- workflow() %>%
  add_recipe(recipeMg) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Mg_metrics <- Mg_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Mg2_evaluation <- workflow() %>%
  add_recipe(recipeMg2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Mg2_metrics <- Mg2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(Mg_metrics)
print(Mg2_metrics)
```

Now for Mn...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Mn_evaluation <- workflow() %>%
  add_recipe(recipeMn) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Mn_metrics <- Mn_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Mn2_evaluation <- workflow() %>%
  add_recipe(recipeMn2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Mn2_metrics <- Mn2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(Mn_metrics)
print(Mn2_metrics)
```

Now for S...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Sul_evaluation <- workflow() %>%
  add_recipe(recipeSul) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Sul_metrics <- Sul_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Sul2_evaluation <- workflow() %>%
  add_recipe(recipeSul2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Sul2_metrics <- Sul2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(Sul_metrics)
print(Sul2_metrics)
```

Now for Si...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Si_evaluation <- workflow() %>%
  add_recipe(recipeSi) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Si_metrics <- Si_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Si2_evaluation <- workflow() %>%
  add_recipe(recipeSi2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Si2_metrics <- Si2_evaluation %>%
  collect_metrics()

#Print the RMSE and R-square for both models
print(Si_metrics)
print(Si2_metrics)
```

Now for Ti...

```{r}
###### MODEL 1 #######
# Fit the model using cross-validation
Ti_evaluation <- workflow() %>%
  add_recipe(recipeTi) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

# Extract the results of the model
Ti_metrics <- Ti_evaluation %>%
  collect_metrics()

###### MODEL 2 ######
# Fit the model using cross-validation
Ti2_evaluation <- workflow() %>%
  add_recipe(recipeTi2) %>%
  add_model(gamma_spec) %>%
  fit_resamples(resamples = samples)

>>>>

---
title: "Data analysis code"
author: Erick E. Mollinedo
date: '`r format(Sys.Date(), "%B %d, %Y")`'
format: html
editor: visual
---

@@

```{r}
library(here)
library(readxl)
library(tidyverse)
library(corrplot)
library(grid)
library(RColorBrewer)
library(ggpubr)
library(rstatix)
library(tidymodels)
```

@@

### Summarize Concentrations

The following chunk of code is to create the descriptive statistics tables for all chemical species concentrations. The tables will be saves as RDS files, so later they can be loaded as tables using `gt()`.

```{r}
#Create data frames for both LPG and Biomass groups
biomass <- concentration %>% filter(fueltype == "Biomass") %>% select(-c(arm, filter_id))
lpg <- concentration %>% filter(fueltype == "LPG") %>% select(-c(arm, filter_id))

#Format dfs longer
biomass_long <- biomass %>% #Biomass dataframe
  pivot_longer(cols = -fueltype,
               names_to = "Species", values_to = "Concentration")

lpg_long <- lpg %>% #LPG dataframe
  pivot_longer(cols = -fueltype,
               names_to = "Species", values_to = "Concentration")

#Create summary tables for each group
biomass_summary <- biomass_long %>% #Biomass fuel
  group_by(Species) %>% 
  summarise(N = n(),
            Mean = round(mean(Concentration), 2),
            SD = round(sd(Concentration), 2),
            Median = round(median(Concentration), 2),
            IQR = round(IQR(Concentration), 2),
            Min = round(min(Concentration), 2),
            Max = round(max(Concentration), 2))

lpg_summary <- lpg_long %>% #LPG fuel
  group_by(Species) %>% 
  summarise(N = n(),
            Mean = round(mean(Concentration), 2),
            SD = round(sd(Concentration), 2),
            Median = round(median(Concentration), 2),
            IQR = round(IQR(Concentration), 2),
            Min = round(min(Concentration), 2),
            Max = round(max(Concentration), 2))

#Save dataframes in RDS format so they can later be uploaded as tables
write_rds(biomass_summary, here("results", "tables", "biomass-summary.rds"))
write_rds(lpg_summary, here("results", "tables", "lpg-summary.rds"))
```

### Correlations

To answer the first question (correlations by species), I will plot the correlations in a correlation plot using the `corrplot()` function. I used the previous code, since it was already developed, but editing to add more detail to the plot.

@@

### Fuel type contrasts

To answer question 2 (species differences among type of fuel), I will conduct a T-test using the `tidymodels` package

@@

  tibble( #Format the tibble
    variable = variable,
    statistic = t_test$statistic,
    p_value = t_test$p.value,
    estimate = t_test$estimate,
    null_value = t_test$null.value,
    alternative = t_test$alternative,
    method = t_test$method,
    conf_low = t_test$conf.int[1],
    conf_high = t_test$conf.int[2])
})

@@

To see a graphical representation of this test, check the boxplots generated in the `data_exploring_code.qmd` file.

### PM2.5 and BC Modeling

To answer questions 3 and 4 (modeling between PM2.5, BC and other chemical species), I will conduct a regression analysis. But first I will merge the data from `hapin_filters` with the `concentration` data. Modeling can be difficult if there are missing values, so I will additionally clean the `hapin_merged` df, so it removes all NAs.

@@

#Remove bc and co, and additionally drop all NAs
hapin_merged <- hapin_merged %>%
  select(-c(bc, co)) %>%
  drop_na()

@@

Now conducting the model selection using the `tidymodels` package. For the models, I tried two multivariate models, first, using the gamma distribution family, since the concentrations of each chemical species could be explained by the gamma distribution. And the second model, using a simple linear regression. Also, the data was split into training (80%) and test data (20%).

```{r}
# Split the data into training and testing sets
rngseed = 1234
set.seed(rngseed) # for reproducibility
data_split <- initial_split(hapin_merged, prop = 0.8)
train_data <- training(data_split)
test_data <- testing(data_split)
```

Now, creating the formulas for each question. Also, computing the model specifications, the workflow and fitting the models

```{r}
#Create formulas
pm25formula <- pm25 ~ Mg + Al + Si + S + K + Ca + Ti + Mn + Fe + Zn + BC
bcformula <- BC ~ smoke + coil + trash + kerosene + smoky + crop + stove_other + fueltype

#Set model specifications
lin_mod <- linear_reg() %>% set_engine("lm") #Linear
glm_mod <- linear_reg(mode = "regression") %>% #GLM gamma
  set_engine("glm", family = Gamma(link = "log"))

## ----Model 1: PM2.5 Linear---- ##
linpm_wflow <- workflow() %>% #Workflow
	add_model(lin_mod) %>% 
	add_formula(pm25formula)

linpm_fit <- linpm_wflow %>% fit(data = train_data) #Fit the model

## ----Model 2: PM2.5 GLM Gamma---- ##
glmpm_wflow <- workflow() %>% #Workflow
  add_model(glm_mod) %>% 
  add_formula(pm25formula)

glmpm_fit <- glmpm_wflow %>% fit(data = train_data) #Fit the model

## ----Model 3: BC Linear---- ##
linbc_wflow <- workflow() %>% #Workflow
	add_model(lin_mod) %>% 
	add_formula(bcformula)

linbc_fit <- linbc_wflow %>% fit(data = train_data) #Fit the model

## ----Model 4: BC GLM Gamma---- ##
glmbc_wflow <- workflow() %>% #Workflow
  add_model(glm_mod) %>% 
  add_formula(bcformula)

glmbc_fit <- glmbc_wflow %>% fit(data = train_data) #Fit the model
```

Extracting the results from each model.

```{r}
# Extract and print the results of the models
results_pmlin <- linpm_fit %>% #PM2.5 Linear
  extract_fit_parsnip() %>%
  tidy()

results_pmglm <- glmpm_fit %>% #PM2.5 GLM
  extract_fit_parsnip() %>%
  tidy()

results_bclin <- linbc_fit %>% #BC Linear
  extract_fit_parsnip() %>%
  tidy()

results_bcglm <- glmbc_fit %>% #BC GLM
  extract_fit_parsnip() %>%
  tidy()

#Print the results
print(results_pmlin)
print(results_pmglm)
print(results_bclin)
print(results_bcglm)
```

Now, doing an evaluation of each model First, by estimating the metrics (In this case I will use RMSE, R-squared and MAE). First computing the predictions to make comparisons with the observed values.

```{r}
#Compute the predictions
#PM2.5 models
lin_pm_pred <- linpm_fit %>% predict(train_data)
glm_pm_pred <- glmpm_fit %>% predict(train_data)
#BC models
lin_bc_pred <- linbc_fit %>% predict(train_data)
glm_bc_pred <- glmbc_fit %>% predict(train_data)

#Compute the metrics
#PM2.5 models
lin_pm_metrics <-  bind_cols(train_data, lin_pm_pred) %>% metrics(truth = pm25, estimate = .pred) 
glm_pm_metrics <- bind_cols(train_data, glm_pm_pred) %>% metrics(truth = pm25, estimate = .pred)
#BC Models
lin_bc_metrics <-  bind_cols(train_data, lin_bc_pred) %>% metrics(truth = BC, estimate = .pred) 
glm_bc_metrics <- bind_cols(train_data, glm_bc_pred) %>% metrics(truth = BC, estimate = .pred)

#Print the metrics
print(lin_pm_metrics)
print(glm_pm_metrics)
print(lin_bc_metrics)
print(glm_bc_metrics)
```

Continue the model evaluation, plotting the observed vs predicted values for each model.

```{r}
#Create dataframes to compute the plots
#PM2.5
pred_pm1 <- data.frame(predicted = lin_pm_pred$.pred, model = "linear")
pred_pm2 <- data.frame(predicted = glm_pm_pred$.pred, model = "GLM")
#BC
pred_bc1 <- data.frame(predicted = lin_bc_pred$.pred, model = "linear")
pred_bc2 <- data.frame(predicted = glm_bc_pred$.pred, model = "GLM")

#Merge data frames
#PM2.5
pmdata_plot <- bind_rows(pred_pm1, pred_pm2) %>% 
  mutate(observed = rep(train_data$pm25, 2))
#BC
bcdata_plot <- bind_rows(pred_bc1, pred_bc2) %>% 
  mutate(observed = rep(train_data$BC, 2))

#Create pred-obs plot

#Define colors
my_colors <- c("GLM" = "#E1BE6A", "linear" = "#40B0A6")

#PM2.5
pmpreds <- ggplot(pmdata_plot) +
  geom_point(aes(x = observed, y = predicted, color= model), size= 2) +
  facet_wrap(~ model)+
  scale_color_manual(values = my_colors)+
  labs(x = "Observed", y = "Predicted") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
  theme_bw()

#BC
bcpreds <- ggplot(bcdata_plot) +
  geom_point(aes(x = observed, y = predicted, color= model), size= 2) +
  facet_wrap(~ model)+
  scale_color_manual(values = my_colors)+
  labs(x = "Observed", y = "Predicted") +
  theme_bw()

#Visualize plots
pmpreds
bcpreds

#Save plots
ggsave(here("results", "figures", "obs-pred_pm25.png"), plot = pmpreds, width = 10, height = 6, dpi = 300)
ggsave(here("results", "figures", "obs-pred_bc.png"), plot = bcpreds, width = 10, height = 6, dpi = 300)
```

Also, computing the residuals for each model and plotting them.

```{r}
#Calculate residuals and create data frame
#PM2.5
pmdata_plot <- pmdata_plot %>% mutate(observation = 1:988, residuals = observed - predicted)
#BC
bcdata_plot <- bcdata_plot %>% mutate(observation = 1:988, residuals = observed - predicted)

# Plot residuals using ggplot
#PM2.5
pmresiduals <- ggplot(pmdata_plot) +
  geom_point(aes(x = observation, y = residuals, color= model), size= 2) +
  geom_abline(intercept = 0, slope = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = my_colors)+
  labs(x = "Data point",
       y = "Residuals")+
  theme_bw()

#BC
bcresiduals <- ggplot(bcdata_plot) +
  geom_point(aes(x = observation, y = residuals, color= model), size= 2) +
  geom_abline(intercept = 0, slope = 0, linetype = "dashed", color = "black") +
  scale_color_manual(values = my_colors)+
  labs(x = "Data point",
       y = "Residuals")+
  theme_bw()

#Visualize plots
pmresiduals
bcresiduals

#Save plots
ggsave(here("results", "figures", "residuals-pm25.png"), plot = pmresiduals, width = 10, height = 6, dpi = 300)
ggsave(here("results", "figures", "residuals-bc.png"), plot = bcresiduals, width = 10, height = 6, dpi = 300)
```

And then computing the 90% confidence intervals for the predicted values for each model.

```{r}
## ----Model 1: PM2.5 Linear---- ##
# Compute prediction intervals
pred_int_linpm <- predict(linpm_fit, train_data, type = "conf_int", level = 0.90) %>%
  mutate(observation = 1:494)

# Combine the lower and upper bounds with the observed values
pred_int_linpm <- cbind(lin_pm_pred, pred_int_linpm, train_data$pm25)
pred_int_linpm <- pred_int_linpm %>% rename(observed = "train_data$pm25")
```

Finally, plotting the predictions and the prediction intervals.

```{r}
#Prediction interval plot for the linear model of PM2.5
ggplot(pred_int_linpm) +
	geom_errorbar(aes(x = observed, ymin = .pred_lower, ymax = .pred_upper), width = 25) +
	geom_point(aes(x = observed, y = .pred, color = ".pred"),	shape = 5) +
	labs(x = "Observed", y = "Predicted", title = "Predicted observation with Prediction Intervals") +
	geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
	scale_color_manual(values = my_colors) +
	theme_minimal()
```

This plot looks crowded for the linear model, so I won't use it as part of the final assessment of this and the other models. However I will keep it for future references.

Then, doing a final evaluation of the model using the test data. I decided to select the linear multivariate models for both models (PM2.5 and BC).

```{r}
## ----PM2.5 Linear---- ##
#Fit the model using the test data
linpm_fit_test <- linpm_wflow %>% fit(data = test_data)

#Make predictions in the test data
pm_preds2 <- linpm_fit_test %>% predict(test_data)

#Compute the metrics
linpm_metrics_test <- bind_cols(test_data, pm_preds2) %>% metrics(truth = pm25, estimate = .pred)

print(linpm_metrics_test)

## ----BC Linear---- ##
#Fit the model using the test data
linbc_fit_test <- linbc_wflow %>% fit(data = test_data)

#Make predictions in the test data
bc_preds2 <- linbc_fit_test %>% predict(test_data)

#Compute the metrics
linbc_metrics_test <- bind_cols(test_data, bc_preds2) %>% metrics(truth = BC, estimate = .pred)

print(linbc_metrics_test)
```

### Save outputs as tables

Creating a table to save all the metric values in one table. First creating new dataframes that includes all metrics for the PM2.5 and BC models.

```{r}
#Bind dfs that contain all RMSEs and R-squares from the PM2.5 and BC models
metricspm_final <- bind_rows(lin_pm_metrics, glm_pm_metrics)
metricsbc_final <- bind_rows(lin_bc_metrics, glm_bc_metrics)

#Delete non-necessary columns
metricspm_final <- metricspm_final %>% select(-.estimator)
metricsbc_final <- metricsbc_final %>% select(-.estimator)

# Creating a vector of observation types
obs <- rep(c(rep("M1", 3), rep("M2", 3)), length.out = 6)

# Adding the 'obs' variable to 'metrics_final'
metricspm_final$obs <- obs
metricsbc_final$obs <- obs
```
